```{r}
#| label: set-up-results
#| echo: false
#| message: false
library(targets)
library(tarchetypes)
library(ggplot2)
library(gt)
```

# Results

### Standardized functioning

The effect of plant functional group removal, summer temperature and annual precipitation on single ecosystem functions.

```{r}
#| label: fig-function
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 10
#| fig-cap: Standardized function in response to number of functional groups present across different precipitation levels and alpine, sub-alpine and boreal habitats. The different functions are shown in different colours and the different groups have different symbols.

tar_read(function_figure)

```


### Average multifunctionality

The number of functional groups present and temperature both increased average multifunctionality.

```{r}
#| label: fig-multifunctionality
#| echo: false
#| warning: false
#| fig-cap: Average multifuntionality in response to number of functional groups present and summer temperature (habitat). The colours represent the different habitats. The points are the raw data and the lines represent the predictions from the model (not that the line is a simple lm model for now!).

tar_read(multifunctionality_figure)

```

```{r}
#| label: tbl-model-multi
#| echo: false
#| warning: false
#| tbl-cap: Model output from testing effect of number of functional groups present, summer temperature and annual precipitation on multifunctionality.

tar_read(model_multi_nr_out)

```

The identity of the functional groups removed had an effect on average multifunctionality.
Multifunctionality was reduced when no plants, only forbs or only bryophytes were present.


```{r}
#| label: fig-multifunctionality2
#| echo: false
#| warning: false
#| fig-cap: Average multifuntionality in response to number of functional groups present across different precipitation levels and alpine, sub-alpine and boreal habitats. The different functional groups present are shown in different colours.

tar_read(multifunctionality_figure2)

```

```{r}
#| label: tbl-model-multi-trt
#| echo: false
#| warning: false
#| tbl-cap: Model output from testing effect of treatment on multifunctionality.

tar_read(model_multi_trt_out)

```

### Factorial approach: presence of forbs, graminoids and bryophytes

Predicted multifunctionality from the factorial model (forb × gram × bryo × temperature × precipitation) for each functional group's presence (0 = absent, 1 = present), stratified by habitat (summer temperature) or precipitation.

```{r}
#| label: fig-factorial-multifun-temp
#| echo: false
#| warning: false
#| fig-cap: Predicted multifunctionality in response to presence of forbs, graminoids or bryophytes, by habitat (summer temperature). Points show predictions at each site's observed climate; lines connect mean predictions at absence (0) and presence (1).

tar_read(factorial_multifun_temp_figure)

```

```{r}
#| label: fig-factorial-multifun-prec
#| echo: false
#| warning: false
#| fig-cap: Predicted multifunctionality in response to presence of forbs, graminoids or bryophytes, by precipitation. Points show predictions at each site's observed climate; lines connect mean predictions at absence (0) and presence (1).

tar_read(factorial_multifun_prec_figure)

```

```{r}
#| label: tbl-model-multi-factorial
#| echo: false
#| warning: false
#| tbl-cap: Model output from testing effect of treatment (factorial design), summer temperature and annual precipitation on multifunctionality.

tar_read(model_multi_fact_out)
  
```
